Classification of ordinal data in deep learning: an experimental study

### 2019-03-12 paper/ieee.tex
Mejoras ieee. Glogit

% *** MATH PACKAGES ***
%
\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

@@

\usepackage{ulem}

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract or keywords.
	\begin{abstract}
		The abstract goes here.
	\end{abstract}

	% Note that keywords are not normally used for peerreview papers.
	\begin{IEEEkeywords}
		IEEE, IEEEtran, journal, \LaTeX, paper, template.
	\end{IEEEkeywords}

@@

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\textcolor{red}{Habría que restructurar la introducción: 1) deep learning y CNN, pero corto 2) clasificación ordinal, modelos de umbral, POM y funciones de enlace (más bien breve) 3) clasificación ordinal profunda 4) objetivo del paper (un estudio experimental sobre funciones de enlace y otros parámetros en deep learning ordinal). learning rate y batch normalization, si se mete, debería ser más breve}

	1) \IEEEPARstart{D}{eep learning}, introduced by Yann Lecun~\cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy protection~\cite{yu2017iprivacy}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly}. \textcolor{blue}{\sout{Deep learning discovers complex structures in large datasets by using the backpropagation algorithm to change the internal parameters of the model and compute better representations in each layer from the features in the previous layer.} Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that comes in the form of multiple arrays.} CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. On image classification tasks, each colour channel is represented by a 2D array. In this case, convolutional layers extract the main features from the pixels of the images and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. These outputs are compared against the correct values.

	2) Ordinal classification problems are those classification tasks where labels are ordered, and there are different inter-classes importances for each pair of classes. This kind of problem can be treated as a nominal classification problem, but this discards ordinal information. A better approach is to use specific methods that take into account this kind of information to improve the performance of the classification model. \textcolor{blue}{\sout{One way to use the ordinal information is to evaluate the model using an ordinal metric. Multiple metrics exist in the literature of machine learning and statistics \cite{cruz2014metrics,mehdiyev2016evaluating}. Kappa index is a well-known statistic coefficient defined by Cohen \cite{cohen1960coefficient} to measure inter-rater agreement on classifying elements into a set of categories. Later, Weighted Kappa (WK) is a modified version of the Kappa statistic calculated to allow assigning different weights to different levels of aggregation between two variables. An ordinal weighted kappa loss was described in a previous work~\cite{de2018weighted}, which is a differentiable cost function based on the WK metric. This previous work proved that using these functions improves the model performance and reduce the overfitting risk. Different weights can be assigned in an ordinal classification problem. In a linear penalization, the weight is proportional to the distance between the predicted and the real class (in number of categories). In the Quadratic Weighted Kappa (QWK) the penalization is proportional to the square of the distance. In this work, we will use the QWK metric and the QWK loss.} The Proportional Odds Model (POM)~\cite{agresti2010analysis} is an ordinal alternative to the softmax function that was designed for ordinal regression. It belongs to a wider family of models called Cumulative Link Models (CLM). It is inspired in the concept of a latent variable that is projected in an n-dimensional space and a set of thresholds that divides this space into the different ordinal levels. This kind of models uses a link function which can be of different types. The most common link function is the \texttt{logit} one, that is used in POM. However, there are other functions that can be explored. This model will be described in depth in Section \ref{sect:ordinalproblem}.}

	\textcolor{blue}{3) Ordinal models can also be applied to deep learning models. In the case of convolutional networks, the model projection used by the threshold model will be obtained from the last layer of the network. When working with a 1-dimensional space, the last layer will have only one neuron, and its value will be used to classify every sample in the corresponding class according to the thresholds. Some previous works have used the POM in traditional neural networks~\cite{gutierrez2016ordinal}, but it has not been applied to convolutional networks yet. Also, there are some link functions that has not been explored.}

	\textcolor{blue}{4) In this paper, an experimental study regarding link functions for Cumulative Link Models will be made. Also, other parameters that can affect the training process and the model performance, like the learning rate of the optimization algorithm and the batch size, and its interaction will be studied. The nominal version of this model will be used as a baseline for comparison. We will contrast the results obtained with statistical analysis to provide more robust conclusions. An approximated ANOVA III \cite{miller1997beyond} test followed by a posthoc Tukey's test \cite{tukey1949comparing} will be performed because of the limitations of the computational time required to run a higher number of executions.}

@@

%	\textcolor{red}{The backpropagation algorithm is a stochastic gradient descent algorithm that minimises a predefined loss function. It has been used in the last years in many works~\cite{leonard1990improvement,yu1995dynamic,krizhevsky2012imagenet,de2018weighted} for training shallow and deep neural networks. It updates the weights of each layer after backpropagating the loss function gradients through the network. The learning rate hyper-parameter controls the strength of the changes that are applied to the weights. Some works have checked the importance of finding the optimal value for this parameter~\cite{senior2013empirical} and have tried different approaches to improve the training process~\cite{smith2017cyclical}.}
%	
%	\textcolor{red}{Batch normalization is another technique used for this kind of networks. It reduces the internal covariate shift by normalizing layer inputs. It was presented in 2015~\cite{ioffe2015batch} and gives a critical enhancement during the training phase. Batch size is another important hyper-parameter that should be adjusted as it affects the weight updates and also the normalization~\cite{keskar2016large}\cite{radiuk2017impact}.}
%	
%	The softmax function is widely used for the output layer in neural networks for classification tasks. It is a simple and efficient function for multi-class problems, but it is not the best when working with ordinal data. In this paper, the Cumulative Link Model (CLM)~\cite{gutierrez2016ordinal,agresti2010analysis} will be used instead of the softmax function. Different link functions will be explored to compare their performance. Also, the influence of other parameters like the learning rate of the optimization algorithm and the batch size will be studied. We will contrast the results obtained with statistical analysis to provide more robust conclusions. An approximated ANOVA III \cite{miller1997beyond} test followed by a posthoc Tukey's test \cite{tukey1949comparing} will be performed because of the limitations of the computational time required to run a higher number of executions.

@@

	J. de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the QWK loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function across three different databases, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning. First, they defined QWK metric as follows:
	\begin{equation}
	\textcolor{blue}{\text{QWK} = 1 - \frac{\sum\limits_{i,j} \omega_{i,j} O_{i,j}}{\sum\limits_{i,j} \omega_{i,j} E_{i,j}},}
	\end{equation}
	\textcolor{blue}{where $\omega$ is the penalization matrix (in this case, quadratic weights are considered), $O$ is the confusion matrix and $E$ is the normalized outer product between the prediction and the true vector.}

	\textcolor{blue}{Then, they provided the QWK loss ($\text{QWK}_l$) definition below:}
	\begin{equation}
	\textcolor{blue}{\text{QWK}_l = \frac{\sum\limits_{k=1}^N \sum\limits_{q=1}^Q \omega_{t_k, q} P(y = \mathcal{C}_q | \mathbf{x}_k)}{\sum\limits_{i=1}^Q \frac{N_i}{N} \sum\limits_{j=1}^Q ( \omega_{i,j} \sum\limits_{k=1}^N P(y = \mathcal{C}_j | \mathbf{x}_k))},}
	\end{equation}
	\textcolor{blue}{where $\text{QWK}_l \in [0,2]$, $\mathbf{x}_k$ and $t_k$ are the input data and the real class of the $k$-th sample, $Q$ is the number of classes, $N$ is the number of samples, $N_i$ is the number of samples of the $i$th class, $P(y = \mathcal{C}_q | \mathbf{x}_k)$ is the probability that the $k$th sample belongs to class $\mathcal{C}_q$ and $\omega_{i,j}$ are the elements of the penalization matrix. Generally, $\omega_{i,j} = \frac{|i-j|^n}{(C-1)^n}$, where $\omega_{i,j} \in [0,1]$. In this case, the $\text{QWK}_l$ is a function to be minimized while the QWK metric must be maximized.}

	\textcolor{blue}{Z. Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using convolutional neural networks. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively resolve these classification sub-problems, taking into account the correlation between them.}

	Christopher Beckham and Christopher Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. They evaluated this approach in the context of deep learning on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results.	\textcolor{blue}{Also, they proposed a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.}

@@

	\textcolor{blue}{H. Li et al. \cite{li2017deep} applied deep learning techniques for solving the ordinal problem of Alzheimer's diagnosis and detecting the different levels of the disease.}

	\textcolor{blue}{Y. Liu et al. \cite{liu2017deep} proposed a new approach of which transforms the ordinal regression problem to binary classification problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationship are extracted automatically.}

	\textcolor{blue}{A. Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature engineering. Also, they applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.}

	\textcolor{blue}{S. Chen et al. \cite{chen2017using} proposed a deep method termed Ranking-CNN. This method combines multiple binary CNNs that are trained with ordinal age labels. The binary outputs are aggregated for the final age prediction and they achieved a tighter error bound for ranking-based age estimation.}

	\textcolor{blue}{H. Fu et al. \cite{fu2018deep} applied deep learning techniques to Monocular Depth Estimation. They introduced a spacing-increasing discretization strategy to treat the problem as an ordinal regression problem. They improved the performance when training the network with an ordinary regression loss. Also, they used a multi-scale network structure that avoids unnecessary spatial pooling.}

	\textcolor{blue}{Y. Liu et al. \cite{liu2018constrained} proposed a constrained optimization formulation for the ordinal regression problem which minimizes de negative loglikelihood for multiple categories constrained by the order relationship between instances.}

	\textcolor{blue}{A. Pal et al. \cite{pal2018severity} defined a loss function for CNN that is based on the Earth Mover's Distance and takes into account the ordinal class relationships.}

	\textcolor{blue}{M. ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expresiveness to ordinal scale.}

	\textcolor{blue}{\sout{Currently, there is a great discussion in terms of finding the best activation function, offering good performance and mitigating the effects of the gradient vanishing problem \cite{bengio1994learning,pascanu2013difficulty}. Rectified Linear Unit (ReLU)~\cite{nair2010rectified} is widely used in most deep learning works, but recently, Clevert et al. proposed the Exponential Linear Unit (ELU)~\cite{clevert2015fast}. They proved that ELUs alleviate the vanishing gradient problem via the identity for positive values. In their experiments, ELUs lead not only to faster learning but also to significantly better generalization performance than ReLUs and Leaky ReLU functions (LReLU) on networks with more than five layers. In this way, this paper considers the ELU function for the different experiments.}}

	\textcolor{blue}{\sout{Sergey Ioffe and Christian Szegedy~\cite{ioffe2015batch} described Batch Normalization and its benefits. It reduces the internal covariate shift by normalizing layer inputs. Their method draws its strength from making normalization a part of the model architecture and performing the normalization for each training batch. It allows them to use higher learning rates and be less careful about initialization. It also eliminates the need for using regularization techniques like Dropout.}}

@@

	\textcolor{blue}{The model is inspired by the notion of a latent variable, where $f(\textbf{x})$ represents a one-dimensional mapping obtained from the output of the last layer, which has only one neuron.} The decision rule $r: \mathcal{X} \rightarrow \mathcal{Y}$ is not fitted directly, but stochastic ordering of space $\mathcal{X}$ is satisfied by the following general model form \cite{herbrich2000large}:
	\begin{equation}
		g^{-1}(P(y \preceq \mathcal{C}_q | \mathbf{x})) = b_q - \textcolor{blue}{f(\mathbf{x})}, \quad q = 1, ..., Q-1,
	\end{equation}
	where $g^{-1} : [0,1] \rightarrow (-\infty, +\infty)$ is a monotonic function often termed as the inverse link function, and $b_q$ is the threshold defined for class $\mathcal{C}_q$. Consider the latent variable $y^* = f(\mathbf{x})^* = \textcolor{blue}{f(\mathbf{x})} + \epsilon$, where $\epsilon$ is the random variable of the error. The most common choice for the probability distribution of $\epsilon$ is the logistic function (which is the default function for POM). Label $\mathcal{C}_q$ is predicted if and only if $f(\mathbf{x}) \in [b_{q-1}, b_q]$, where the function $f$ and $\mathbf{b} = (b_0, b_1, ..., b_{Q-1}, b_Q)$ are to be determined from the data. It is assumed that $b_0 = -\infty$ and $b_Q = +\infty$, so the real line defined by $f(\textbf{x}), \textbf{x} \in \mathcal{X}$, is divided into $Q$ consecutive intervals. Each interval corresponds to a category. The constraints $b_1 \le b_2 \le ... \le b_{Q-1}$ ensures that $P(y \preceq \mathcal{C}_q | \mathbf{x})$ increases with $q$ \cite{mccullagh1980regression}.
	\textcolor{blue}{This order is achieved by defining the first threshold and calculating the rest of thresholds from the first in the following form:}

	\begin{equation}
	\textcolor{blue}{b_n = b_1 + \sum_{i=1}^{n-1} \alpha_i^2, \quad n = 2, ..., N},
	\end{equation}
	\textcolor{blue}{where $b_1$ and $\alpha_i$ are learnable parameters, and $N$ is the number of classes.}

@@

	These three types of links are explained below and represented in Figure \ref{fig:linkfunctions}. They all follow the same form $P(y \preceq \mathcal{C}_q | \mathbf{x}) = \Phi(b_q - \textcolor{blue}{f(\mathbf{x})})$ for a continuous cdf~$\Phi$.

	\begin{figure}[!t]
		\centering
		\includegraphics[width=3in]{img/linkfunctions.png}
		\caption{Representation of link functions.}
		\label{fig:linkfunctions}
	\end{figure}

	\begin{itemize}
		\item \texttt{Logit}. \texttt{logit} link function is the function used for the Proportional Odds Model. The \texttt{logit} link is:
		\begin{equation}
		\begin{aligned}
		\text{logit}[P(y \preceq \mathcal{C}_q | \mathbf{x})] = \log\frac{P(y \preceq \mathcal{C}_q | \mathbf{x})}{1 - P(y \preceq \mathcal{C}_q | \mathbf{x})}=& \\ = b_q - \textcolor{blue}{f(\mathbf{x})}, \quad q = 1, ..., Q-1,
		\end{aligned}
		\label{eq:logit}
		\end{equation}		
		or the equivalent expression:		
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \frac{1}{1 + e^{-(b - \textcolor{blue}{f(\mathbf{x})})}}.
		\label{eq:logit2}
		\end{equation}

		\item \texttt{Probit}. \texttt{probit} link function is the inverse of the standard normal cumulative distribution function (cdf) $\Phi$. Its expression is:
		\begin{equation}
		\begin{aligned}
		\Phi^{-1}[P(y \preceq \mathcal{C}_q | \mathbf{x})] = b_q - \textcolor{blue}{f(\mathbf{x})}, \quad &q = 1, ..., Q-1,\\
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \Phi(b_q - \textcolor{blue}{f(\mathbf{x})}), \quad &q = 1, ..., Q-1,
		\end{aligned}
		\label{eq:probit}
		\end{equation}		
		which can also be expressed as:
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \int_{-\infty}^{b_q - \textcolor{blue}{f(\mathbf{x})}} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x.
		\label{eq:probit2}
		\end{equation}

		\item \textit{Complementary log-log}. Like the \texttt{logit} and the \texttt{probit} transformation, the complementary log-log transformation takes a response that is restricted to the $(0,1)$ interval and converts it into something in the $(-\infty, +\infty)$ interval. Complementary log-log expression is:
		\begin{equation}
		\begin{aligned}
		\log[-\log[1 - P(y \preceq \mathcal{C}_q | \mathbf{x})]] =&\\
		= b_q - \textcolor{blue}{f(\mathbf{x})}, \quad q = 1, ..., Q-1,&
		\end{aligned}
		\label{eq:cloglog}
		\end{equation}
		that is:
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = 1 - e^{-e^{b_q - \textcolor{blue}{f(\mathbf{x})}}}, \quad q = 1, ..., Q-1.
		\label{eq:cloglog2}
		\end{equation}
	\end{itemize}

@@

	\subsection{Model}
	\label{sect:model}
	CNNs have been used for both datasets. The different architectures of CNN used in these experiments are presented in tables \ref{table:CNNArchitecture} and \ref{table:ResNetArchitecture}. The architecture for DR is the same that was used in \cite{de2018weighted} and the network for Adience is a small Residual Network (ResNet) \cite{he2016deep} that was used in \cite{beckham2017unimodal}. \textcolor{blue}{The most important parameters for convolutional layers are the number of filters that are used to make the convolution operation, the size of these filters and the stride, which is the number of pixels that the filter is moved for obtaining the next pixel. Pooling layers have got similar parameters: the pool size is the number of pixels that will be involved in the pooling operation, and the stride represents the same concept that in convolutional layers.} For convolutional layers, ConvWxH@FsS~=~F filters of size WxH and stride S. For pooling layers, PoolWxHsS~=~pool size of WxH and stride S.

>>>>

% *** MATH PACKAGES ***
%
\usepackage{amsmath}
\usepackage{mathrsfs}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

@@

%\usepackage{ulem}

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract or keywords.
	\begin{abstract}
		This paper proposes a deep convolutional neural network model for ordinal regression based on the use of a family of probabilistic ordinal link functions in the output layer called Cumulative Link Models. This kind of models projects each pattern into a 1-dimensional space. In this case, the projections are estimated by a non-linear deep neural network. A set of ordered thresholds splits this space into the different classes of the problem. Different link functions will be studied in this experimental study, and the results will be contrasted with statistical analysis. To further improve the results, we combine these ordinal models with a loss function that takes into account the distance between the categories, based on the weighted Kappa index. The experiments run over two different ordinal classification problems, and the statistical tests confirm that these models improve the results of a nominal model and outperform the proposals considered in the literature.
	\end{abstract}

	% Note that keywords are not normally used for peerreview papers.
	\begin{IEEEkeywords}
		Deep learning, Ordinal Regression, Link models.
	\end{IEEEkeywords}

@@

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\IEEEPARstart{D}{eep learning}, introduced by Yann Lecun~\cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy protection~\cite{yu2017iprivacy}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that comes in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. On image classification tasks, each colour channel is represented by a 2D array. In this case, convolutional layers extract the main features from the pixels of the images and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. These outputs are compared against the correct values.

	Ordinal classification problems are those classification tasks where labels are ordered, and there are different inter-classes importances for each pair of classes. This kind of problem can be treated as a nominal classification problem, but this discards ordinal information. A better approach is to use specific methods that take into account this kind of information to improve the performance of the classification model. The Proportional Odds Model (POM)~\cite{agresti2010analysis} is an ordinal alternative to the softmax function that was designed for ordinal regression. It belongs to a wider family of models called Cumulative Link Models (CLM). It is inspired in the concept of a latent variable that is projected in an n-dimensional space and a set of thresholds that divides this space into the different ordinal levels. This kind of models uses a link function which can be of different types. The most common link function is the \texttt{logit} one, that is used in POM. However, there are other functions that can be explored. This model will be described in depth in Section \ref{sect:ordinalproblem}.

	Ordinal models can also be applied to deep learning models. In the case of convolutional networks, the model projection used by the threshold model will be obtained from the last layer of the network. When working with a 1-dimensional space, the last layer will have only one neuron, and its value will be used to classify every sample in the corresponding class according to the thresholds. Some previous works have used the POM in traditional neural networks~\cite{gutierrez2016ordinal}, but it has not been applied to convolutional networks yet. Also, there are some link functions that has not been explored.

	In this paper, an experimental study regarding link functions for Cumulative Link Models will be made. Also, other parameters that can affect the training process and the model performance, like the learning rate of the optimization algorithm and the batch size, and its interaction will be studied. The nominal version of this model will be used as a baseline for comparison. We will contrast the results obtained with statistical analysis to provide more robust conclusions. An approximated ANOVA III \cite{miller1997beyond} test followed by a posthoc Tukey's test \cite{tukey1949comparing} will be performed because of the limitations of the computational time required to run a higher number of executions.

@@

@@

	J. de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the QWK loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function across three different databases, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning. First, they defined QWK metric as follows:
	\begin{equation}
	\text{QWK} = 1 - \frac{\sum\limits_{i,j} \omega_{i,j} O_{i,j}}{\sum\limits_{i,j} \omega_{i,j} E_{i,j}},
	\end{equation}
	where $\omega$ is the penalization matrix (in this case, quadratic weights are considered), $O$ is the confusion matrix and $E$ is the normalized outer product between the prediction and the true vector.

	Then, they provided a continuous version of the QWK metric ($\text{QWK}_c$) based on the probabilities of the predictions:
	\begin{equation}
	\text{QWK}_c = 1 - \frac{\sum\limits_{k=1}^N \sum\limits_{q=1}^Q \omega_{t_k, q} P(y = \mathcal{C}_q | \mathbf{x}_k)}{\sum\limits_{i=1}^Q \frac{N_i}{N} \sum\limits_{j=1}^Q ( \omega_{i,j} \sum\limits_{k=1}^N P(y = \mathcal{C}_j | \mathbf{x}_k))},
	\end{equation}
	where $\text{QWK}_c \in [-1,1]$, $\mathbf{x}_k$ and $t_k$ are the input data and the real class of the $k$-th sample, $Q$ is the number of classes, $N$ is the number of samples, $N_i$ is the number of samples of the $i$th class, $P(y = \mathcal{C}_q | \mathbf{x}_k)$ is the probability that the $k$th sample belongs to class $\mathcal{C}_q$ and $\omega_{i,j}$ are the elements of the penalization matrix. Generally, $\omega_{i,j} = \frac{|i-j|^n}{(C-1)^n}$, where $\omega_{i,j} \in [0,1]$.

	Then, they defined the loss function based on the QWK as follows:
	\begin{equation}
	\mathscr{L} = \log(1 - \text{QWK}_c),
	\end{equation}
	where $\mathscr{L} \in [-\infty, \log 2]$. This loss is a function to be minimized while the QWK metric must be maximized.

	Z. Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using convolutional neural networks. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively resolve these classification sub-problems, taking into account the correlation between them.

	C. Beckham and C. Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. They evaluated this approach in the context of deep learning on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results.	Also, they proposed a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.

@@

	H. Li et al. \cite{li2017deep} applied deep learning techniques for solving the ordinal problem of Alzheimer's diagnosis and detecting the different levels of the disease.

	Y. Liu et al. \cite{liu2017deep} proposed a new approach of which transforms the ordinal regression problem to binary classification problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationship are extracted automatically.

	A. Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature engineering. Also, they applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.

	S. Chen et al. \cite{chen2017using} proposed a deep method termed Ranking-CNN. This method combines multiple binary CNNs that are trained with ordinal age labels. The binary outputs are aggregated for the final age prediction and they achieved a tighter error bound for ranking-based age estimation.

	H. Fu et al. \cite{fu2018deep} applied deep learning techniques to Monocular Depth Estimation. They introduced a spacing-increasing discretization strategy to treat the problem as an ordinal regression problem. They improved the performance when training the network with an ordinary regression loss. Also, they used a multi-scale network structure that avoids unnecessary spatial pooling.

	Y. Liu et al. \cite{liu2018constrained} proposed a constrained optimization formulation for the ordinal regression problem which minimizes the negative loglikelihood for multiple categories constrained by the order relationship between instances.

	A. Pal et al. \cite{pal2018severity} defined a loss function for CNN that is based on the Earth Mover's Distance and takes into account the ordinal class relationships.

	M. ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expresiveness to ordinal scale.

@@

	The model is inspired by the notion of a latent variable, where $f(\textbf{x})$ represents a one-dimensional mapping obtained from the output of the last layer, which has only one neuron. The decision rule $r: \mathcal{X} \rightarrow \mathcal{Y}$ is not fitted directly, but stochastic ordering of space $\mathcal{X}$ is satisfied by the following general model form \cite{herbrich2000large}:
	\begin{equation}
		g^{-1}(P(y \preceq \mathcal{C}_q | \mathbf{x})) = b_q - f(\mathbf{x}), \quad q = 1, ..., Q-1,
	\end{equation}
	where $g^{-1} : [0,1] \rightarrow (-\infty, +\infty)$ is a monotonic function often termed as the inverse link function, and $b_q$ is the threshold defined for class $\mathcal{C}_q$. Consider the latent variable $y^* = f(\mathbf{x})^* = f(\mathbf{x}) + \epsilon$, where $\epsilon$ is the random variable of the error. The most common choice for the probability distribution of $\epsilon$ is the logistic function (which is the default function for POM). Label $\mathcal{C}_q$ is predicted if and only if $f(\mathbf{x}) \in [b_{q-1}, b_q]$, where the function $f$ and $\mathbf{b} = (b_0, b_1, ..., b_{Q-1}, b_Q)$ are to be determined from the data. It is assumed that $b_0 = -\infty$ and $b_Q = +\infty$, so the real line defined by $f(\textbf{x}), \textbf{x} \in \mathcal{X}$, is divided into $Q$ consecutive intervals. Each interval corresponds to a category. The constraints $b_1 \le b_2 \le ... \le b_{Q-1}$ ensures that $P(y \preceq \mathcal{C}_q | \mathbf{x})$ increases with $q$ \cite{mccullagh1980regression}.
	This order is achieved by defining the first threshold and calculating the rest of thresholds from the first in the following form:

	\begin{equation}
	b_n = b_1 + \sum_{i=1}^{n-1} \alpha_i^2, \quad n = 2, ..., N,
	\end{equation}
	where $b_1$ and $\alpha_i$ are learnable parameters, and $N$ is the number of classes.

@@

	These three types of links are explained below and represented in Figure \ref{fig:linkfunctions}. They all follow the same form $P(y \preceq \mathcal{C}_q | \mathbf{x}) = \Phi(b_q - f(\mathbf{x}))$ for a continuous cdf~$\Phi$.

	\begin{figure}[!t]
		\centering
		\includegraphics[width=3in]{img/linkfunctions.pdf}
		\caption{Representation of link functions.}
		\label{fig:linkfunctions}
	\end{figure}

	\begin{itemize}
		\item \texttt{Logit}. \texttt{logit} link function is the function used for the Proportional Odds Model. The \texttt{logit} link is:
		\begin{equation}
		\begin{aligned}
		\text{logit}[P(y \preceq \mathcal{C}_q | \mathbf{x})] = \log\frac{P(y \preceq \mathcal{C}_q | \mathbf{x})}{1 - P(y \preceq \mathcal{C}_q | \mathbf{x})}=& \\ = b_q - f(\mathbf{x}), \quad q = 1, ..., Q-1,
		\end{aligned}
		\label{eq:logit}
		\end{equation}		
		or the equivalent expression:		
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \frac{1}{1 + e^{-(b - f(\mathbf{x}))}}.
		\label{eq:logit2}
		\end{equation}

		\item \texttt{Probit}. \texttt{probit} link function is the inverse of the standard normal cumulative distribution function (cdf) $\Phi$. Its expression is:
		\begin{equation}
		\begin{aligned}
		\Phi^{-1}[P(y \preceq \mathcal{C}_q | \mathbf{x})] = b_q - f(\mathbf{x}), \quad &q = 1, ..., Q-1,\\
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \Phi(b_q - f(\mathbf{x})), \quad &q = 1, ..., Q-1,
		\end{aligned}
		\label{eq:probit}
		\end{equation}		
		which can also be expressed as:
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = \int_{-\infty}^{b_q - f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x.
		\label{eq:probit2}
		\end{equation}

		\item \textit{Complementary log-log}. Like the \texttt{logit} and the \texttt{probit} transformation, the complementary log-log transformation takes a response that is restricted to the $(0,1)$ interval and converts it into something in the $(-\infty, +\infty)$ interval. Complementary log-log expression is:
		\begin{equation}
		\begin{aligned}
		\log[-\log[1 - P(y \preceq \mathcal{C}_q | \mathbf{x})]] =&\\
		= b_q - f(\mathbf{x}), \quad q = 1, ..., Q-1,&
		\end{aligned}
		\label{eq:cloglog}
		\end{equation}
		that is:
		\begin{equation}
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = 1 - e^{-e^{b_q - f(\mathbf{x})}}, \quad q = 1, ..., Q-1.
		\label{eq:cloglog2}
		\end{equation}
	\end{itemize}

@@

	\subsection{Model}
	\label{sect:model}
	CNNs have been used for both datasets. The different architectures of CNN used in these experiments are presented in tables \ref{table:CNNArchitecture} and \ref{table:ResNetArchitecture}. The architecture for DR is the same that was used in \cite{de2018weighted} and the network for Adience is a small Residual Network (ResNet) \cite{he2016deep} that was used in \cite{beckham2017unimodal}. The most important parameters for convolutional layers are the number of filters that are used to make the convolution operation, the size of these filters and the stride, which is the number of pixels that the filter is moved for obtaining the next pixel. Pooling layers have got similar parameters: the pool size is the number of pixels that will be involved in the pooling operation, and the stride represents the same concept that in convolutional layers. For convolutional layers, ConvWxH@FsS~=~F filters of size WxH and stride S. For pooling layers, PoolWxHsS~=~pool size of WxH and stride S.
